---
title: "The anatomy lab: locating sentiment in a language model"
format: html
engine: knitr
vignette: >
  %\VignetteIndexEntry{The anatomy lab: locating sentiment in a language model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# The heavy steps need a local GGUF model and glmnet; a checker (CRAN,
# r-universe) has neither. We gate those chunks on availability and fall back to
# an illustrative curve, so this vignette always renders with a figure. Point
# REBIRTH_DEMO_MODEL (or REBIRTH_TEST_MODEL_QWEN) at a GGUF file to render the
# figure from real activations.
model_path <- Sys.getenv("REBIRTH_DEMO_MODEL", Sys.getenv("REBIRTH_TEST_MODEL_QWEN", ""))
have_model <- nzchar(model_path) && file.exists(model_path)
have_glmnet <- requireNamespace("glmnet", quietly = TRUE)
run_real <- have_model && have_glmnet
```

`llm_trace()` observes the residual stream during a forward pass and hands it back
as a plain data frame. That turns a classic mechanistic-interpretability question
--- *where in the network does a concept become linearly readable?* --- into an
ordinary R modelling exercise. Here the concept is **sentiment**.

The recipe is deliberately classical: trace a fixed contrast set, take a concept
direction with `prcomp()`, fit one cross-validated linear probe per layer, and
plot decodability against depth. Then we *steer* along that direction and check
the effect on held-out prompts.

## AUC needs no dependency

Decodability is the probe's AUC. AUC is the rank-based Mann--Whitney statistic ---
twelve lines of base R, with average ranks giving exact tie handling --- so we
inline it rather than take a package dependency.

```{r}
#| label: auc
demo_auc <- function(scores, labels) {
  pos <- as.numeric(scores)[labels]
  neg <- as.numeric(scores)[!labels]
  n_pos <- length(pos)
  n_neg <- length(neg)
  r <- rank(c(pos, neg)) # average ranks => correct tie handling
  (sum(r[seq_along(pos)]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

# Stratified percentile-bootstrap CI (resample within class, fixed seed).
demo_auc_ci <- function(scores, labels, B = 2000L, seed = 1L) {
  scores <- as.numeric(scores)
  pos <- scores[labels]
  neg <- scores[!labels]
  est <- demo_auc(scores, labels)
  set.seed(seed)
  boot <- vapply(seq_len(B), function(i) {
    demo_auc(c(sample(pos, replace = TRUE), sample(neg, replace = TRUE)),
             c(rep(TRUE, length(pos)), rep(FALSE, length(neg))))
  }, numeric(1))
  ci <- stats::quantile(boot, c(0.025, 0.975), names = FALSE)
  c(auc = est, lower = ci[1L], upper = ci[2L])
}

# Sanity checks: perfect separation -> 1, label flip -> 1 - AUC, a tie case.
stopifnot(
  all.equal(demo_auc(c(3, 4, 5, 0, 1, 2), c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)), 1),
  all.equal(demo_auc(c(1, 2, 1, 3), c(TRUE, TRUE, FALSE, FALSE)), 0.375)
)
```

## A fixed contrast set

We use a fixed, committed set of opposite-sentiment sentences (no cherry-picking).
A handful, for illustration:

```{r}
#| label: prompts
pos <- c(
  "I absolutely loved this film; it was a masterpiece.",
  "The service was outstanding and the staff were welcoming.",
  "What a wonderful day -- everything went perfectly."
)
neg <- c(
  "I hated this film; it was a boring waste of time.",
  "The service was appalling and the staff were rude.",
  "What a miserable day -- everything went wrong."
)
train <- c(pos, neg)
y <- c(rep(TRUE, length(pos)), rep(FALSE, length(neg)))
```

The full runnable demo (a larger contrast set, held-out prompts, and the steering
check) is `tests/demos/demo-A-anatomy-lab.R` in the package repository.

## Trace, probe, and the money plot

With a model loaded, one `llm_trace()` call captures the last-token residual at
every layer; then for each layer we fit a **ridge** logistic probe with
`glmnet::cv.glmnet()` (ridge because the hidden width far exceeds the number of
prompts) and read its honest out-of-fold AUC.

```{r}
#| label: probe-real
#| eval: !expr run_real
library(rebirth)
m <- llm(model_path)
tr <- llm_trace(m, train, positions = "last", components = "residual")

layer_matrix <- function(tr, L) {
  x <- as.matrix(tr, layer = L, component = "residual")
  x[order(as.integer(sub("[.].*$", "", rownames(x)))), , drop = FALSE]
}
yf <- factor(ifelse(y, "pos", "neg"), levels = c("neg", "pos"))
set.seed(1)
foldid <- sample(rep(seq_len(5), length.out = length(y)))

band <- sort(unique(tr$layer))
auc <- lower <- upper <- numeric(length(band))
for (i in seq_along(band)) {
  cvfit <- glmnet::cv.glmnet(layer_matrix(tr, band[i]), yf, family = "binomial",
                             alpha = 0, foldid = foldid, type.measure = "auc",
                             keep = TRUE)
  oof <- cvfit$fit.preval[, match(cvfit$lambda.min, cvfit$lambda)]
  ci <- demo_auc_ci(oof, y)
  auc[i] <- ci[["auc"]]; lower[i] <- ci[["lower"]]; upper[i] <- ci[["upper"]]
}
res <- list(layers = band, auc = auc, lower = lower, upper = upper)
close(m)
```

```{r}
#| label: probe-illustrative
#| eval: !expr isFALSE(run_real)
# Illustrative curve (no model available at render time): sentiment becomes
# linearly readable in the middle layers. Run the chunk above on your own model
# to reproduce this figure from real activations.
L <- 1:24
set.seed(1)
auc <- pmin(pmax(0.5 + 0.47 / (1 + exp(-(L - 7) / 1.6)) + rnorm(24, 0, 0.01), 0.5), 0.99)
w <- (0.5 + 0.5 * (1 - auc)) * 0.09
res <- list(layers = L, auc = auc, lower = pmax(auc - w, 0), upper = pmin(auc + w, 1))
```

```{r}
#| label: money-plot
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Probe AUC by layer, rising from chance to near one in the middle layers."
L <- res$layers
op <- par(mar = c(4.4, 4.4, 3, 1), mgp = c(2.6, 0.7, 0), las = 1)
plot(L, res$auc, type = "n", ylim = c(0.45, 1),
     xlab = "Transformer layer", ylab = "Probe AUC (out-of-fold)",
     main = sprintf("Where sentiment becomes linearly readable (%s)",
                    if (run_real) "real" else "illustrative"))
abline(h = 0.5, lty = 2, col = "grey45")
arrows(L, res$lower, L, res$upper, angle = 90, code = 3, length = 0.03, col = "grey30")
lines(L, res$auc, col = "#3B6EA5", lwd = 1.5)
points(L, res$auc, pch = 21, bg = "#3B6EA5", col = "white", cex = 1.2)
b <- L[which.max(res$auc)]
abline(v = b, lty = 3, col = adjustcolor("#D1495B", 0.4))
points(b, max(res$auc), pch = 21, bg = "#D1495B", col = "white", cex = 1.6)
text(max(L), 0.455, sprintf("best: layer %d (AUC %.2f)", b, max(res$auc)),
     adj = c(1, 0), font = 2, col = "#D1495B")
par(op)
```

The AUC rises from chance in the early layers to a plateau in the middle of the
network: that is *where sentiment lives* for this model.

## Steering along the direction

`llm_steer()` returns a **new** model handle that adds `coef * direction` to the
residual stream at one layer; the original handle is untouched. A concept
direction comes straight from `prcomp()` of the traced activations. Reading the
effect at a **later** layer on held-out prompts shows it propagates forward ---
a genuine downstream shift, not the injected vector read straight back.

```{r}
#| label: steer
#| eval: false
pc <- prcomp(layer_matrix(tr, b), center = TRUE, scale. = FALSE)
direction <- pc$rotation[, 1]
if (demo_auc(pc$x[, 1], y) < 0.5) direction <- -direction

m_pos <- llm_steer(m, layer = b, direction = direction, coef = 6)
llm_generate(m,     "In one sentence, describe your morning.", seed = 1)
llm_generate(m_pos, "In one sentence, describe your morning.", seed = 1)
close(m_pos)
```

The full script adds the quantitative check: a one-sided paired Wilcoxon test on
the held-out prompts' downstream sentiment read-out, positive vs. negative
steering. Everything here is base-R idiom over plain data frames and matrices ---
the model is just another object in your session.
