---
title: "The anatomy lab: locating sentiment in a language model"
format: html
engine: knitr
vignette: >
  %\VignetteIndexEntry{The anatomy lab: locating sentiment in a language model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# The heavy steps need a local GGUF model and glmnet; a checker (CRAN,
# r-universe) has neither. We gate those chunks on availability and fall back to
# an illustrative curve, so this vignette always renders with a figure. Point
# RELM_DEMO_MODEL (or RELM_TEST_MODEL_QWEN) at a GGUF file to render the
# figure from real activations.
model_path <- Sys.getenv("RELM_DEMO_MODEL", Sys.getenv("RELM_TEST_MODEL_QWEN", ""))
have_model <- nzchar(model_path) && file.exists(model_path)
have_glmnet <- requireNamespace("glmnet", quietly = TRUE)
run_real <- have_model && have_glmnet
```

`llm_trace()` observes the residual stream during a forward pass and hands it back
as a plain data frame. That turns a classic mechanistic-interpretability question
--- *where in the network does a concept become linearly readable?* --- into an
ordinary R modelling exercise. Here the concept is **sentiment**.

The recipe is deliberately classical: trace a fixed contrast set, take a concept
direction with `prcomp()`, fit one cross-validated linear probe per layer, and
plot decodability against depth. Then we *steer* along that direction and check
the effect on held-out prompts.

## AUC needs no dependency

Decodability is the probe's AUC. AUC is the rank-based Mann--Whitney statistic ---
twelve lines of base R, with average ranks giving exact tie handling --- so we
inline it rather than take a package dependency.

```{r}
#| label: auc
demo_auc <- function(scores, labels) {
  pos <- as.numeric(scores)[labels]
  neg <- as.numeric(scores)[!labels]
  n_pos <- length(pos)
  n_neg <- length(neg)
  r <- rank(c(pos, neg)) # average ranks => correct tie handling
  (sum(r[seq_along(pos)]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

# Stratified percentile-bootstrap CI (resample within class, fixed seed).
demo_auc_ci <- function(scores, labels, B = 2000L, seed = 1L) {
  scores <- as.numeric(scores)
  pos <- scores[labels]
  neg <- scores[!labels]
  est <- demo_auc(scores, labels)
  set.seed(seed)
  boot <- vapply(seq_len(B), function(i) {
    demo_auc(c(sample(pos, replace = TRUE), sample(neg, replace = TRUE)),
             c(rep(TRUE, length(pos)), rep(FALSE, length(neg))))
  }, numeric(1))
  ci <- stats::quantile(boot, c(0.025, 0.975), names = FALSE)
  c(auc = est, lower = ci[1L], upper = ci[2L])
}

# Sanity checks: perfect separation -> 1, label flip -> 1 - AUC, a tie case.
stopifnot(
  all.equal(demo_auc(c(3, 4, 5, 0, 1, 2), c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)), 1),
  all.equal(demo_auc(c(1, 2, 1, 3), c(TRUE, TRUE, FALSE, FALSE)), 0.375)
)
```

## A fixed contrast set

We use a fixed, committed set of opposite-sentiment sentences (no cherry-picking).
A handful, for illustration:

```{r}
#| label: prompts
pos <- c(
  "I absolutely loved this film; it was a masterpiece.",
  "The service was outstanding and the staff were welcoming.",
  "What a wonderful day -- everything went perfectly."
)
neg <- c(
  "I hated this film; it was a boring waste of time.",
  "The service was appalling and the staff were rude.",
  "What a miserable day -- everything went wrong."
)
train <- c(pos, neg)
y <- c(rep(TRUE, length(pos)), rep(FALSE, length(neg)))
```

The full runnable demo (a larger contrast set, held-out prompts, and the steering
check) is `tests/demos/demo-A-anatomy-lab.R` in the package repository.

## Trace, probe, and the money plot

With a model loaded, one `llm_trace()` call captures the last-token residual at
every layer; then for each layer we fit a **ridge** logistic probe with
`glmnet::cv.glmnet()` (ridge because the hidden width far exceeds the number of
prompts) and read its honest out-of-fold AUC.

```{r}
#| label: probe-real
#| eval: !expr run_real
library(relm)
m <- llm(model_path)
tr <- llm_trace(m, train, positions = "last", components = "residual")

layer_matrix <- function(tr, L) {
  x <- as.matrix(tr, layer = L, component = "residual")
  x[order(as.integer(sub("[.].*$", "", rownames(x)))), , drop = FALSE]
}
yf <- factor(ifelse(y, "pos", "neg"), levels = c("neg", "pos"))
set.seed(1)
foldid <- sample(rep(seq_len(5), length.out = length(y)))

band <- sort(unique(tr$layer))
auc <- lower <- upper <- numeric(length(band))
for (i in seq_along(band)) {
  cvfit <- glmnet::cv.glmnet(layer_matrix(tr, band[i]), yf, family = "binomial",
                             alpha = 0, foldid = foldid, type.measure = "auc",
                             keep = TRUE)
  oof <- cvfit$fit.preval[, match(cvfit$lambda.min, cvfit$lambda)]
  ci <- demo_auc_ci(oof, y)
  auc[i] <- ci[["auc"]]; lower[i] <- ci[["lower"]]; upper[i] <- ci[["upper"]]
}
res <- list(layers = band, auc = auc, lower = lower, upper = upper)
close(m)
```

```{r}
#| label: probe-illustrative
#| eval: !expr isFALSE(run_real)
# Illustrative curve (no model available at render time): sentiment becomes
# linearly readable in the middle layers. Run the chunk above on your own model
# to reproduce this figure from real activations.
L <- 1:24
set.seed(1)
auc <- pmin(pmax(0.5 + 0.47 / (1 + exp(-(L - 7) / 1.6)) + rnorm(24, 0, 0.01), 0.5), 0.99)
w <- (0.5 + 0.5 * (1 - auc)) * 0.09
res <- list(layers = L, auc = auc, lower = pmax(auc - w, 0), upper = pmin(auc + w, 1))
```

```{r}
#| label: money-plot
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Probe AUC by layer, rising from chance to near one in the middle layers."
L <- res$layers
op <- par(mar = c(4.4, 4.4, 3, 1), mgp = c(2.6, 0.7, 0), las = 1)
plot(L, res$auc, type = "n", ylim = c(0.45, 1),
     xlab = "Transformer layer", ylab = "Probe AUC (out-of-fold)",
     main = sprintf("Where sentiment becomes linearly readable (%s)",
                    if (run_real) "real" else "illustrative"))
abline(h = 0.5, lty = 2, col = "grey45")
arrows(L, res$lower, L, res$upper, angle = 90, code = 3, length = 0.03, col = "grey30")
lines(L, res$auc, col = "#3B6EA5", lwd = 1.5)
points(L, res$auc, pch = 21, bg = "#3B6EA5", col = "white", cex = 1.2)
b <- L[which.max(res$auc)]
abline(v = b, lty = 3, col = adjustcolor("#D1495B", 0.4))
points(b, max(res$auc), pch = 21, bg = "#D1495B", col = "white", cex = 1.6)
text(max(L), 0.455, sprintf("best: layer %d (AUC %.2f)", b, max(res$auc)),
     adj = c(1, 0), font = 2, col = "#D1495B")
par(op)
```

The AUC rises from chance in the early layers to a plateau in the middle of the
network: that is *where sentiment lives* for this model.

## Steering along the direction

`llm_steer()` returns a **new** model handle that adds `coef * direction` to the
residual stream at one layer; the original handle is untouched. A concept
direction comes straight from `prcomp()` of the traced activations. Reading the
effect at a **later** layer on held-out prompts shows it propagates forward ---
a genuine downstream shift, not the injected vector read straight back.

```{r}
#| label: steer
#| eval: false
pc <- prcomp(layer_matrix(tr, b), center = TRUE, scale. = FALSE)
direction <- pc$rotation[, 1]
if (demo_auc(pc$x[, 1], y) < 0.5) direction <- -direction

m_pos <- llm_steer(m, layer = b, direction = direction, coef = 6)
llm_generate(m,     "In one sentence, describe your morning.", seed = 1)
llm_generate(m_pos, "In one sentence, describe your morning.", seed = 1)
close(m_pos)
```

The full script adds the quantitative check: a one-sided paired Wilcoxon test on
the held-out prompts' downstream sentiment read-out, positive vs. negative
steering. Everything here is base-R idiom over plain data frames and matrices ---
the model is just another object in your session.

## Reading is not causing

The money plot found where sentiment is *readable*. It is tempting to conclude that
those readable units are the ones the model *uses* for sentiment. They are not --- and
seeing the gap is worth a figure, because it is the representation-space cousin of
"correlation is not causation".

`llm_ablate()` returns a handle with a set of residual units pinned to zero;
`llm_logits()` reads the next-token distribution before and after. We rank the units
of the best layer three ways and measure the mean next-token KL from the un-ablated
model, each against a size-matched **random** control averaged over several draws:

- **load-bearing units** --- the top-*k* by residual RMS: the outlier / "massive
  activation" dimensions that carry most of the residual norm and are load-bearing for
  next-token prediction on *any* prompt;
- **concept-readout units** --- the top-*k* by |probe coefficient|: exactly the units
  the linear probe reads sentiment from;
- **matched random-*k*** --- the control.

```{r}
#| label: decodability-vs-causality
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Mean next-token KL vs neurons ablated, log scale: load-bearing units far above a matched-random control, while concept-readout units track random."
# Illustrative of the measured RELATIVE pattern (the absolute KL scale varies with
# model and layer -- e.g. impact KL ~ O(1) on Qwen2.5-0.5B, ~ O(0.03) on Gemma 4 E4B
# -- but the orders-of-magnitude separation does not). Reproduce it from your own
# model with tests/demos/demo-A-anatomy-lab.R: run_demo_A(extended = TRUE) writes the
# real three-series figure (demoA-A4-decodability.png).
ks <- c(1, 2, 4, 8, 16, 32, 64)
load_bearing    <- c(0.05, 0.10, 1.20, 2.00, 3.00, 3.50, 4.00) # top-k by residual RMS
concept_readout <- c(1e-4, 2e-4, 5e-4, 1.6e-3, 3e-3, 5e-3, 1.5e-2) # top-k by |probe coef|
matched_random  <- c(4e-4, 7e-4, 2e-3, 6e-3, 1.2e-2, 3e-2, 5e-2) # size-matched control

xx <- log2(ks)
cols <- c(load = "#D1495B", random = "grey45", concept = "#3B6EA5")
op <- par(mar = c(4.4, 4.6, 3, 1), mgp = c(2.7, 0.7, 0), las = 1)
plot(xx, load_bearing, type = "n", log = "y", xaxt = "n",
     ylim = c(1e-4, 6), xlab = "neurons ablated (k)",
     ylab = "Mean next-token KL (log)",
     main = "Reading is not causing: readable units are not the causal locus")
axis(1, at = xx, labels = ks)
abline(h = 10^(-4:0), col = "grey93")
for (s in list(list(matched_random, "random"), list(concept_readout, "concept"),
               list(load_bearing, "load"))) {
  lines(xx, s[[1]], col = cols[[s[[2]]]], lwd = 1.6)
  points(xx, s[[1]], pch = 21, bg = cols[[s[[2]]]], col = "white", cex = 1.3)
}
legend("topleft", bty = "n", cex = 0.9, pch = 21, lwd = 1.6, pt.bg = cols, col = cols,
       legend = c("load-bearing units (top-k residual RMS)",
                  "matched random-k (control)",
                  "concept-readout units (top-k |probe coef|)"))
par(op)
```

Ablating the load-bearing units dwarfs random by one to three orders of magnitude:
intervention is extremely **unit-specific** --- *which* units you remove is everything.
But the concept-readout units --- where sentiment is most *readable* --- are **no more
disruptive than matched random**. Decodability is not the causal locus.

### Sufficiency is not necessity

Does this contradict the steering result just above? No, and the reconciliation is the
whole point. Steering **adds** `coef * direction` and demonstrably moves sentiment: the
readable *direction* is causally **sufficient**. Ablation **pins to zero** the
direction's top basis *units* and changes almost nothing: those units are not
**necessary**. Sufficiency is not necessity --- different operation (add vs. pin-to-zero),
different object (a distributed direction vs. a handful of coordinate units), different
outcome metric (a concept shift vs. a concept-blind KL). The concept is *distributed*,
so no small set of readout coordinates is individually load-bearing. A linear probe
tells you where information is **readable**, not which units the computation **needs**.

### A magnitude caveat

Be careful about the "concept-readout below random" reading in the figure. The KL here
is **concept-blind** --- it measures total next-token disruption, not sentiment --- and
zero-ablating a unit perturbs the residual by roughly its own activation magnitude. The
concept-readout units happen to be low-magnitude, while the random draws are only
*count*-matched (not *magnitude*-matched) and so carry a larger typical magnitude. The
honest claim is therefore "**no more disruptive than matched random**", never "below
random"; a magnitude-matched control would be needed to say more.

### Where unit-level causal importance lives

Ranking units by *measured* causal effect --- ablate a unit (or group), read the KL,
repeat --- is exactly what activation patching and a probe-with-interventions do. That
is Phase-4 `llm_probe()` territory; the pattern already exists in the package's own WP5
honesty fixture (`calibrate_kl` in `intervene_kl.rs`), which selected its targeted
neurons by a per-neuron ablation-KL scan rather than by any linear readout.
