---
title: "See, compare and record model interventions"
format: html
engine: knitr
vignette: >
  %\VignetteIndexEntry{See, compare and record model interventions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

The development graphics interface separates three questions: where an operation
is configured, what was observed under two recorded inputs, and which coefficients
a worker actually applied over time. It uses base graphics and ordinary data
frames. It never runs inference implicitly or changes model weights.

## A download-free numeric example

The following editable records are **synthetic**, not model measurements. Their
four coordinate differences were independently specified as `0.5, -1, 1, 0`.
Real observations come directly from `llm_generate(on_state = ...)`.

```{r}
#| fig-width: 10
#| fig-height: 7
library(relm)
source(system.file("examples", "graphics-states.R", package = "relm"), local = TRUE)
comparison <- llm_compare(reference, intervention, context, layer = 2)
stopifnot(identical(comparison$difference, c(0.5, -1, 1, 0)))
plot(comparison, main = "Synthetic observed-state comparison")
```

Reference and intervention use the same vertical scale. The difference panel is
signed intervention minus reference. Attributes contain sampled-token IDs,
applied steering, source positions and the union of the two top-k distributions.
A token missing from one truncated top-k table is unknown, not probability zero;
these tables cannot establish full-distribution KL or answer quality.

## Record a bounded applied history

```{r}
#| fig-width: 10
#| fig-height: 7
history <- llm_timeline(reference, max_states = 4)
for (i in 2:6) {
  state <- example_state(as.integer(i), coefficient = 0,
                         revision = 1L, after = 1L)
  history <- llm_timeline(state, history, max_states = 4)
}
plot(history, main = "Synthetic applied steering; truncated window")
history[, c("state_id", "coef", "applied_after_state", "effective_source_pos")]
```

This history retains states 3–6. The update after state 1 was effective at state 2;
truncation must not invent an update at state 3. The audit annotation retains the
original boundary even when the marker is outside the visible window. A sampled
token is not necessarily committed output: cancellation and stop removal can
prevent it from appearing in final text. Returning a coefficient to zero removes
its future contribution, but does not erase prior KV history.

`max_states` bounds whole observations; `max_bytes` bounds the helper's admitted
materialization. Large histories evict oldest complete states. A single state
that cannot fit raises `relm_error_oom`. Holding many old histories yourself or
retaining all activation states is outside this working budget.

## Collect real observations without retaining every activation

The model example below is not executed during vignette rendering. Supply an
already downloaded supported GGUF, its independently recorded SHA-256 and the
actual build revision. Use exactly the same settings for the paired runs.
`llm_tokens()` is not a substitute for the generation template's prompt IDs:
its special-token flags differ. `llm_compare()` uses recorded matching model,
settings, prompt text and token count, plus **every prior generated token ID**.
The record is checked for consistency, not authenticated.

```{r}
#| eval: false
m <- llm("/path/to/model.gguf", backend = "cpu", context_length = 512)
changed <- llm_steer(m, layer = 2L, direction = direction, coef = 0.5)
plot(changed, layers = c(1:4, 12))

settings <- list(seed = 17, chat = FALSE, temperature = 0, top_p = 0.95,
  max_tokens = 32L, stop = NULL, context_length = 512L, backend = "cpu",
  relm_version = as.character(packageVersion("relm")),
  engine_revision = "your recorded build revision")

# Run sequentially, waiting for each promise before starting the next model job.
# Save only the selected state's activations, all sampled IDs (<= 1024),
# and a bounded audit table. Keep callbacks short.
run <- function(model, prompt) {
  result <- new.env(parent = emptyenv())
  result$ids <- integer(); result$state <- NULL; result$history <- NULL
  result$promise <- llm_generate(model, prompt, chat = settings$chat,
    max_tokens = settings$max_tokens, seed = settings$seed,
    temperature = settings$temperature, top_p = settings$top_p,
    stop = settings$stop, async = TRUE, layers = 2L, top = 5L,
    on_state = function(state) {
      result$ids <- c(result$ids, state$step$token_id)
      result$history <- llm_timeline(state, result$history, max_states = 32L)
      if (state$step$state_id == 1L) {
        result$state <- state
        result$prefix <- result$ids
      }
      NULL  # history is data, not a steering reply
    })
  result
}
```

After each promise settles, pair the captured states using matching records:

```{r}
#| eval: false
record <- function(run) list(model_sha256 = recorded_sha256,
  settings = settings, generated_tokens = run$prefix)
paired <- llm_compare(baseline$state, treated$state,
  context = list(reference = record(baseline), intervention = record(treated)),
  layer = 2L)
plot(paired)
plot(treated$history)
close(changed); close(m)
```

At state k, the current sampled token has not yet entered the input history for
that state. Divergence in any earlier generated ID withholds numerical differences
and labels the result descriptive. A single aligned comparison describes one
controlled input, not general causal efficacy. Evaluate on held-out prompts,
repeated seeds and external outcome measures before claiming a useful direction.

## Interpret the map and export your results

`plot(model)` shows configured entries, not coefficients that a live worker is
currently applying. Detailed llama/qwen2 maps label attention, MLP and residual
sites; Qwen2 attention-output capture is unavailable and is labelled accordingly.
Other architectures use generic blocks. Gaps and configured steering/ablation are
explicit. The diagram is schematic metadata, not a graph inferred from correlation.
Static component projections are marked `P` at their configured attention or MLP
site. The returned table adds integer `configured_projections` after the existing
columns; `S` and `A` remain residual steering and ablation counts. A zero
projection coefficient still counts as configured. The map does not measure
causal effects or show a worker's changing additive coefficient.

For observations from a projected handle, request that component in generation's
live capture and pass the same component to `llm_compare()`. Prefix-alignment
requirements still apply. Static projection does not create additive steering
revisions: an empty coefficient panel in `llm_timeline()` does not mean that the
handle had no static projection. Label the configured intervention separately.
For more than 32 layers select a subset explicitly; timeline plots admit at most
16 interventions, so select the relevant rows before rendering a larger audit.

All plots return their source tables invisibly and restore graphical parameters.
Scripts own the output device. RStudio uses its usual Plots pane.

```{r}
#| eval: false
pdf("comparison.pdf", width = 10, height = 7)
tryCatch(plot(paired), finally = dev.off())
write.csv(as.data.frame(paired), "comparison.csv", row.names = FALSE)
saveRDS(paired, "comparison-with-provenance.rds")
```

CSV carries columns only; RDS retains the bounded metadata too. Input state traces
may own temporary spill files. A completed comparison contains values rather than
file leases and remains usable after the inputs are released. Reading a missing
or incompatible spill raises a classed error. The reader is intended for
package-produced live Arrow files; its R materialization budget is not a hard
bound on arbitrary hostile IPC decoder allocations or total process memory.
