---
title: "Topic modelling without Python"
format: html
engine: knitr
vignette: >
  %\VignetteIndexEntry{Topic modelling without Python}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# The embedding step needs a local GGUF model; UMAP and HDBSCAN need uwot and
# dbscan. A checker has none of these, so we gate the real pipeline and fall back
# to an illustrative layout keyed off the sample's known topics. Point
# RELM_DEMO_MODEL (or RELM_TEST_MODEL_QWEN) at a GGUF file, and install
# uwot + dbscan, to render the map from real embeddings.
model_path <- Sys.getenv("RELM_DEMO_MODEL", Sys.getenv("RELM_TEST_MODEL_QWEN", ""))
have_model <- nzchar(model_path) && file.exists(model_path)
have_uwot <- requireNamespace("uwot", quietly = TRUE)
have_dbscan <- requireNamespace("dbscan", quietly = TRUE)
run_real <- have_model && have_uwot && have_dbscan
```

A BERTopic-class pipeline --- embed documents, reduce with UMAP, cluster with
HDBSCAN, name the clusters --- is usually a Python affair. With `relm` it is a
handful of base-R lines over CRAN packages the R ecosystem already has, running
fully local with zero Python.

The steps: `llm_embed()` the abstracts, lay them out with `uwot::umap()`, find
clusters with `dbscan::hdbscan()`, and name each cluster by asking the model
itself with `llm_generate()`.

## The corpus

The package ships a small **synthetic** sample so this runs offline; for the real
~5,000-abstract arXiv corpus, use `tests/demos/fetch-abstracts.R`.

```{r}
#| label: data
csv <- system.file("extdata", "abstracts-sample.csv", package = "relm")
abstracts <- if (nzchar(csv)) {
  utils::read.csv(csv, stringsAsFactors = FALSE)
} else {
  # fallback when the vignette is knitted from the source tree
  utils::read.csv(file.path("..", "inst", "extdata", "abstracts-sample.csv"),
                  stringsAsFactors = FALSE)
}
nrow(abstracts)
table(abstracts$label)
```

## Embed, reduce, cluster, name

`set.seed()` plus `n_sgd_threads = 1` makes the UMAP layout reproducible; HDBSCAN
is deterministic given the layout. Cluster names come from the model itself.

```{r}
#| label: pipeline-real
#| eval: !expr run_real
library(relm)
m <- llm(model_path)

emb <- llm_embed(m, abstracts$text, pooling = "mean", normalize = TRUE)

set.seed(20240707)
coords <- uwot::umap(emb, n_neighbors = 15, min_dist = 0.1, n_components = 2,
                     metric = "cosine", n_sgd_threads = 1)
cluster <- dbscan::hdbscan(coords, minPts = 15)$cluster # 0 = noise

# Name each cluster from its most central abstracts.
ids <- sort(setdiff(unique(cluster), 0))
labels <- vapply(ids, function(g) {
  members <- which(cluster == g)
  cen <- colMeans(coords[members, , drop = FALSE])
  medoid <- members[which.min(rowSums(sweep(coords[members, , drop = FALSE], 2, cen)^2))]
  prompt <- paste0(
    "Reply with a SHORT topic label of 2 to 4 words for these abstracts, ",
    "and nothing else.\n\n",
    paste0("- ", abstracts$text[medoid], collapse = "\n"), "\n\nTopic label:"
  )
  trimws(gsub("[\"'`.]", "", llm_generate(m, prompt, max_tokens = 12, temperature = 0, seed = 1)))
}, character(1))
names(labels) <- ids
close(m)
```

```{r}
#| label: pipeline-illustrative
#| eval: !expr isFALSE(run_real)
# Illustrative layout (no model/uwot/dbscan at render time): eight Gaussian blobs
# keyed off the sample's known topics, so the map matches what the real pipeline
# produces. Run the chunk above on your own model for the genuine embedding map.
set.seed(20240707)
labs <- unique(abstracts$label)
ang <- setNames(2 * pi * (seq_along(labs) - 1) / length(labs), labs)
cx <- 6 * cos(ang[abstracts$label]); cy <- 6 * sin(ang[abstracts$label])
coords <- cbind(cx + rnorm(nrow(abstracts), 0, 0.8),
                cy + rnorm(nrow(abstracts), 0, 0.8))
cluster <- as.integer(factor(abstracts$label, levels = labs))
labels <- setNames(labs, seq_along(labs))
# a synthetic embedding (one direction per topic + noise) so the quality and
# structure metrics below have something to read without a model
D <- 24
dirs <- matrix(rnorm(length(labs) * D), length(labs), D)
dirs <- dirs / sqrt(rowSums(dirs^2))
emb <- dirs[cluster, ] + matrix(rnorm(nrow(abstracts) * D, sd = 0.5), nrow(abstracts), D)
emb <- emb / sqrt(rowSums(emb^2))
```

## The labelled map

```{r}
#| label: cluster-map
#| fig-width: 8.5
#| fig-height: 6.5
#| fig-alt: "UMAP scatter of abstracts coloured into topic clusters with topic labels."
ids <- sort(setdiff(unique(cluster), 0))
pal <- adjustcolor(grDevices::hcl.colors(length(ids), "Dark 3"), 0.6)
col <- rep(adjustcolor("grey70", 0.45), length(cluster))
for (j in seq_along(ids)) col[cluster == ids[j]] <- pal[j]

op <- par(mar = c(4.2, 4.2, 3, 1), mgp = c(2.5, 0.7, 0), las = 1)
xr <- range(coords[, 1]); yr <- range(coords[, 2]) # pad so edge labels fit
plot(coords, col = col, pch = 19, cex = 0.5, xlab = "UMAP 1", ylab = "UMAP 2",
     xlim = xr + c(-1, 1) * 0.10 * diff(xr), ylim = yr + c(-1, 1) * 0.06 * diff(yr),
     main = sprintf("Topic modelling without Python (%s)",
                    if (run_real) "real embeddings" else "illustrative"))
halo <- function(x, y, s) {
  o <- 0.006 * diff(par("usr")[1:2])
  for (dx in c(-1, 1)) for (dy in c(-1, 1)) text(x + dx * o, y + dy * o, s, col = "white", font = 2)
  text(x, y, s, font = 2)
}
for (j in seq_along(ids)) {
  members <- which(cluster == ids[j])
  cen <- colMeans(coords[members, , drop = FALSE])
  med <- members[which.min(rowSums(sweep(coords[members, , drop = FALSE], 2, cen)^2))]
  halo(coords[med, 1], coords[med, 2], labels[[as.character(ids[j])]])
}
par(op)
```

Each colour is a discovered topic; grey points are HDBSCAN's noise class. These are
the model's *suggested* groupings, not ground truth --- so the next three steps give
the reader ways to judge them.

## How good are the topics?

A map invites the question *how good are these clusters, really?* Two base-R metrics
answer it, and they disagree in an informative way. **Simplified silhouette** (on the
UMAP coordinates) scores each point by how much closer it sits to its own topic's centre
than to the nearest other topic --- separation in the map you see. **Cohesion** is the
mean cosine of a topic's documents to their centroid in the *original* embedding space,
where UMAP's density distortion cannot flatter the result.

```{r}
#| label: quality
simplified_silhouette <- function(coords, cluster) {
  ids <- sort(setdiff(unique(cluster), 0))
  cen <- t(vapply(ids, function(g) colMeans(coords[cluster == g, , drop = FALSE]),
                  numeric(ncol(coords))))
  d <- vapply(seq_along(ids), function(i) sqrt(rowSums(sweep(coords, 2, cen[i, ])^2)),
              numeric(nrow(coords)))
  sil <- rep(NA_real_, nrow(coords))
  for (i in seq_along(ids)) {
    m <- which(cluster == ids[i]); a <- d[m, i]; b <- apply(d[m, -i, drop = FALSE], 1, min)
    sil[m] <- (b - a) / pmax(a, b)
  }
  sil
}
cohesion <- function(emb, members) {           # = mean cosine of members to the centroid
  U <- emb[members, , drop = FALSE]; U <- U / sqrt(rowSums(U^2))
  sqrt(sum(colMeans(U)^2))
}
ids <- sort(setdiff(unique(cluster), 0))
sil <- simplified_silhouette(coords, cluster)
quality <- data.frame(
  topic = labels[as.character(ids)],
  n = as.integer(table(cluster)[as.character(ids)]),
  silhouette = round(vapply(ids, function(g) mean(sil[cluster == g]), numeric(1)), 2),
  cohesion = round(vapply(ids, function(g) cohesion(emb, which(cluster == g)), numeric(1)), 2)
)
quality[order(-quality$silhouette), ]
```

On real decoder-LLM embeddings the two metrics tell different stories: silhouette
varies from topic to topic (weak, overlapping clusters score lower --- honestly), while
cohesion is *uniformly high* because these embeddings are anisotropic (every document
points into a narrow cone of the space). That gap is exactly why both are reported: the
map can look cleaner than the raw geometry is.

## What is each topic about?

Naming a cluster with the model is one summary; its *distinctive words* are another, and
they are checkable against the abstracts. Ranking by raw frequency would just return
corpus-wide common words. Instead we use the **log-odds-ratio with an informative
Dirichlet prior** (Monroe, Colaresi & Quinn 2008): each word's over-representation in a
topic versus the rest of the corpus, standardised to a z-score, with the prior shrinking
common words toward zero. Twenty lines of base R --- no Python, no topic-model package.

```{r}
#| label: terms
stopwords <- c("the","a","an","and","or","to","of","in","on","for","with","we","our",
               "that","this","is","are","be","by","as","from","using","study","results",
               "method","approach","paper","show","these","which","can","based","present")
tokenize <- function(t) {
  w <- unlist(regmatches(tolower(t), gregexpr("[[:alpha:]]+", tolower(t))))
  w <- w[nchar(w) >= 3]; w[!w %in% stopwords]
}
toks <- lapply(abstracts$text, tokenize)
vocab <- sort(unique(unlist(toks[cluster %in% ids])))
counts <- vapply(ids, function(g)
  as.integer(table(factor(unlist(toks[cluster == g]), levels = vocab))),
  integer(length(vocab)))
rownames(counts) <- vocab                        # vocabulary x topic counts
yw <- rowSums(counts); alpha <- length(vocab) * yw / sum(yw)  # informative prior
a0 <- sum(alpha); N <- sum(counts)
top_terms <- function(gi, n = 8) {
  y <- counts[, gi]; ni <- sum(y); r <- yw - y
  delta <- log((y + alpha) / (ni + a0 - y - alpha)) -
           log((r + alpha) / (N - ni + a0 - r - alpha))
  z <- delta / sqrt(1 / (y + alpha) + 1 / (r + alpha))
  names(sort(z, decreasing = TRUE))[seq_len(n)]
}
data.frame(
  topic = labels[as.character(ids)],
  distinctive_terms = vapply(seq_along(ids),
    function(gi) paste(top_terms(gi), collapse = ", "), character(1))
)
```

The terms discriminate the topics rather than echo the corpus --- and they let a reader
judge each cluster for themselves.

## How do the topics relate?

Finally, which topics are near-duplicates and which stand apart? The cosine similarity
between topic centroids answers it; `hclust()` on `1 - cosine` turns it into a
dendrogram.

```{r}
#| label: structure
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Dendrogram of topics clustered by centroid cosine similarity."
centroids <- t(vapply(ids, function(g) colMeans(emb[cluster == g, , drop = FALSE]),
                      numeric(ncol(emb))))
centroids <- centroids / sqrt(rowSums(centroids^2))
cosm <- pmin(pmax(tcrossprod(centroids), -1), 1)
short <- function(s) ifelse(nchar(s) > 22, paste0(substr(s, 1, 19), "..."), s)
hc <- hclust(as.dist(1 - cosm), method = "average")
hc$labels <- short(labels[as.character(ids)])
op <- par(mar = c(2, 4, 3, 9), mgp = c(2.5, 0.7, 0))
plot(as.dendrogram(hc), horiz = TRUE, xlab = "1 - cosine",
     main = "Which topics are near each other")
par(op)
```

Because decoder-LLM embeddings are anisotropic, every centroid pair sits at a high
absolute cosine; the *relative* structure is what matters, and the demo script draws it
as a heatmap on a data-driven scale. The full runnable version --- the polished map, the
quality panel, per-topic term bars, and the centroid-cosine heatmap --- is
`run_demo_B(extended = TRUE)` in `tests/demos/demo-B-topics.R` (with `fetch-abstracts.R`
for the real ~5,000-abstract arXiv corpus). No Python was involved at any step.
